Characterizing subarctic European vegetation for land surface modeling of ozone pollution and climate risk

### 2021-09-27 VegOzone.tex
Refracture Section 4

\section{Statistical analysis}
\label{sec:stats}
In this section, we assess the climate conditions. To this end, we compute multi-annual mean (referred to as climatology in the following) of environmental key variables (temperature, precipitation, global irradiance) and ground-level \chem{[O_3]} at Svanhovd. For the growing seasons 2018/19, we evaluate the statistical significance of divergences from the norm in these variables (referred to as anomalies). We performed a bias correction accounting for the increase in tropospheric background \chem{[O_3]} and cross-calibrate the derived climatology for Svanvik with long-term observations of \chem{[O_3]} at other sites in northern Fennoscandia as described in \citet{ACPD:Falk2021}.

@@

The derived ozone climatology (\chem{\left<[O_3]\right>}) for Svanvik is shown in Figure~\ref{fig:ozone_climatology_fenoscandic_obs_spline}. Darker colors indicate higher probability to observe these values. On top of the density distributions, a $10\,\unit{days}$ average of daily mean (\chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}}}) is displayed together with $1 \sigma$ uncertainties and SE, respectively. Splines have been fitted through the data for easier inter-comparison. The \chem{[O_3]} density distribution and the \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}}} for Svanvik display a pronounced seasonal cycle. The spring peak occurs slightly earlier ($100\,\unit{doy}$) and amounts to $40\,\unit{ppb}$ while the annual average \chem{\left<[O_3]\right>} is $6.6\,\unit{ppb}$ lower than the Fennoscandic climatology. This is in line with a general decrease in tropospheric ozone with decreasing altitude. The earlier decline in \chem{\left<[O_3]\right>} may point to an influence of coniferous vegetation on the removal of ozone since \chem{CO_2} uptake by coniferous trees has been observed as early as \unit{doy} 100 \citep{TB:Kolari2007, TP:Wallin2013}. The estimated average begin of growing season for deciduous trees ($\left<G_\mathrm{start}\right>_\mathrm{decid}$) is $(150\pm 14)\,\unit{doy}$ at Svanvik (see Section~\ref{subsec:do3se_parameters} for details). In July--September, ozone is occasionally almost completely depleted during nighttime. This hints to ozone uptake by vegetation even at low light intensities during midnight sun conditions in combination with stable planetary boundary layer conditions preventing mixing of ozone rich air. The derived \chem{\left<[O_3]\right>} is likely biased towards lower \chem{[O_3]} due to an increase in northern hemispheric tropospheric background \chem{[O_3]} since the 1990s. To account for this, we derive a bias correction factor by cross-calibrating with the time series taken at Esrange and Pallas (Section~\ref{subsec:anomalies}).

\begin{figure}[t]
  %\includegraphics[width=8.3cm]{ozone_climatology_fenoscandic_obs_spline}
  \includegraphics[width=8.3cm]{fig05}
  \caption{Ozone climatologies. Fennoscandia combines all available data from Jergul/Karasjok, Esrange, and Pallas. The density distributions are shown together with a multi-annual mean of daily \chem{[O_3]} (climatology). Splines have been fitted through daily mean \chem{[O_3]}. The climatology of daily mean ozone is shown together with $1\,\sigma$ standard deviation and standard error of the mean. Since \chem{[O_3]} are strongly correlated for sites at Jergul/Karasjok (NOR), Esrange (SWE), and Pallas (FIN) (see Fig.~\ref{fig:density_distribution}), data of these have been averaged together to derive a climatology for northern Fennoscandia. (a) Northern Fennoscandia; (b) Svanvik.}
  \label{fig:ozone_climatology_fenoscandic_obs_spline}
\end{figure}

\subsection{2018/19 anomalies}
\label{subsec:anomalies}
In the following, we look at the 2018/19 anomalies and discuss the differences between the two years. We also conduct a bias correction for the Svanvik \chem{\left<[O_3]\right>.} As weather extremes like in 2018 are likely to become more frequent with climate change, this will help to assess possible future changes in the vulnerability of subarctic vegetation to ozone.

In Fig.~\ref{fig:anomalies_svanvik}a), temperature anomalies for Svanvik are shown as a percentage of days warmer or colder than $\pm 1\,\sigma$ from the climatological mean for each month. Negative deviations from climatology are displayed as a negative percentage. The annual positive/negative deviations are indicated in the respective corners (right upper/right lower). The hatched area between the dashed lines indicates the expected percentage of values falling above/below $\pm 1\,\sigma$ if a normal distribution is assumed ($15.9\,\unit{\%}$). We find, that the summer of 2018 was significantly warmer than average. Especially in May and July more than $40\,\unit{\%}$ of the days were significantly warmer. Significantly warmer days continued to occur in all months (August--November) except for October. March 2018 had many unusually cold days. The summer of 2019 was fairly average, however, significantly colder days occurred in July, while August had both significantly colder and warmer days. April 2019 had significantly warmer than average days.

@@

Similarly, the precipitation anomalies for Svanvik are displayed in Fig.~\ref{fig:anomalies_svanvik}b). Due to the large interannual variability in observed precipitation, only a few days deviated significantly from the climatology on the $1 \sigma$ level. Therefore, we lowered the $\sigma$ level constraint to $\pm \frac{1}{2}\,\sigma$. The percentage of days wetter/drier than $\pm \frac{1}{2}\,\sigma$ are shown for each month. Positive deviation refers to days wetter than average and negative deviation to drier than average. The hatched area between the dashed lines indicates the expected percentage of values falling above/below $\pm\frac{1}{2}\,\sigma$ if a normal distribution is assumed ($30.8\,\unit{\%}$). Unlike the temperature anomalies, the picture is not as clear. While March 2018 had a significant number of days wetter than unusual, summer and fall (May/July, September/October) had a significant number of days drier than average. 2019 was rather average throughout the year, but with a significant number of days drier than normal in September/October and wetter in December.

The global irradiance anomalies for Svanvik are presented in Fig.~\ref{fig:anomalies_svanvik}c). The percentage of days above or below $\pm 1\,\sigma$ from the climatological mean is displayed for each month. The hatched area between the dashed lines indicates the expected percentage of values falling above/below $\pm 1\,\sigma$ if a normal distribution is assumed. In summer 2018 (May/July), global irradiance showed a significant number of days with higher than average irradiance, while 2019 was generally an average year with a significant number of days darker in spring (April/May).

In summary, 2018 had a significant number of days that were warmer, drier, and brighter than the climatological average, while 2019 was a rather average year.

\textbf{TODO: Remove all content which is part of the technical note}
To assess the significance of 2018/19 anomalies, we estimate the bias due to the change in the tropospheric background \chem{[O_3]} since the 1990s and account for the missing data in June 2018. To reconstruct these missing data, we performed a Reynolds decomposition (Eq.~(\ref{eq:reynolds_decomp})) and cross-calibrate with ozone observations at Esrange and Pallas (companion paper). In the following, we will describe the bias-correction and discuss the significance of \chem{[O_3]} enhancements in 2018.

In 2018, the probability density functions (PDFs) of ozone concentration anomalies (\chem{\Delta[O_3]}) for Esrange and Pallas deviate clearly, but not significantly from the climatological mean (Supplement~Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_residuals}). On average, \chem{[O_3]} had been elevated by $(1.9\pm 5.7)\,\unit{ppb}$ and $(1.9\pm 5.4)\,\unit{ppb}$, respectively. Both resemble normal distributions.

We find a similar skew towards higher \chem{[O_3]} for Svanvik in 2018, while 2019 follows a normal distribution rather well (Supplement~Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_residuals-Svanvik}). In conjunction with the meteorological conditions, we therefore assume that 2019 had been a normal year and the annual average anomaly in 2019 (\chem{\left<\Delta[O_3]\right>}~=~$(1.2\pm 5.0)\,\unit{ppb}$) can be used for the bias correction. We also consult the trends in ozone observations at Esrange and Pallas to strengthen this assumption and compute the average difference between recent years and the mid 1990s (\chem{\Delta_{present-past}\left<[O_3]\right>}). Herein, present is defined as the average over 2010--2018 and past as the time between the respective beginning of data taking until 1997. We do not find a clear trend. For Esrange \chem{\Delta_{present-past}\left<[O_3]\right>} amounts to $(1.5\pm 3.1)\,\unit{ppb}$ and for Pallas to $(-2.5\pm 3.7)\,\unit{ppb}$ \citep[][for comparison]{NILU2003, BER:Ruoho-Airola2015}. Only the observed trend at Esrange is compatible with our assumed bias correction for the Svanvik ozone climatology.

%To further investigate the ozone residuals, we conduct a Student's~t-test ($t = \frac{\chem{\Delta[O_3]}}{\sigma}$), assuming the same sample uncertainty $\sigma$ in both, climatology and the year in question. Results for Esrange and Pallas are shown in Supplement Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_test}. This t-test clearly shows, that at both sites the residuals are significant on the $1\sigma$ level for at least $50\,\unit{\%}$. In Supplement Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_test-Svanvik}, the test statistics for Svanvik are displayed. From Supplement Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_test-Svanvik}a), we can deduce, that the residuals between the Svanvik climatology and the northern Fennoscandia climatology are significant on the $2\sigma$ level for at least half of the data, while the residuals are less certain in 2018 and 2019. In Supplement Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_test-Svanvik}b) the same is shown for the years 2018/19 for the climatology from Svanvik. We find that less than $50\,\unit{\%}$ of ozone concentration residuals are significant on the $1\sigma$ level.

In accordance to the above findings, we adjust the derived ozone climatology for Svanvik by adding a constant bias correction factor $\Delta\chem{[O_3]_\mathrm{bias-corr}} = 1.2\,\unit{ppb}$ and assess the monthly significance of the ozone concentration anomalies in 2018/19. The resulting percentage of days above/below $\pm 1\,\sigma$ from the bias-corrected climatological mean is displayed for each month in Fig.~\ref{fig:anomalies_svanvik}d). Negative deviations from climatology are shown as a negative percentage. Note that missing bars for certain months are both due to data availability and anomalies not meeting the $1\sigma$ criteria. The star indicates results after the reconstruction of missing data in July. In April/May 2018, ozone was significantly enhanced, while no significant enhancement is found in 2019. In both years, ozone concentrations were significantly lower than average in November. These results indicate that the extensive forest fires in Sweden did not contribute to a significant enhancement in \chem{[O_3]} at Svanhovd in 2018, but the stagnant meteorological conditions in April/May in combination with a late start of the growing season. However, there are indications that the sites further south at Esrange and Pallas experienced more days of elevated ozone in 2018 than Svanvik.

%To conclude, ozone concentrations in 2018 have been elevated compared to the respective climatologies, however, on average not significantly by statistical means. At Esrange and Pallas at least $25\,\unit{\%}$ (3.~quantile) of anomalies are significant on a $3\sigma$ level or more, while at Svanvik the 3.~quantile lies at about $2\sigma$. This means that in 2018 the stations at Esrange and Pallas saw more days with elevated ozone than was monitored at Svanvik.

>>>>

\section{Statistical analysis}
\label{sec:stats}
In this section, we assess the climate conditions. To this end, we compute multi-annual mean (referred to as climatology in the following) of environmental key variables (temperature, precipitation, global irradiance) and ground-level \chem{[O_3]} at Svanhovd. For the growing seasons 2018/19, we evaluate the statistical significance of divergences from the norm in these variables (referred to as anomalies). We performed a bias correction accounting for the increase in tropospheric background \chem{[O_3]} and cross-calibrated the derived climatology for Svanvik with long-term observations of \chem{[O_3]} at other sites in northern Fennoscandia as described in \citet{ACPD:Falk2021}.

@@

%Ozone
The derived ozone climatology (\chem{\left<[O_3]\right>}) for Svanvik is shown in Figure~\ref{fig:ozone_climatology_fenoscandic_obs_spline}. Darker colors indicate higher probability to observe these values. On top of the density distributions, a $10\,\unit{days}$ average of daily mean (\chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}}}) is displayed together with $1 \sigma$ uncertainties and SE, respectively. A spline was fitted through the data for comparison. The \chem{[O_3]} density distribution and the \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}}} for Svanvik display a pronounced seasonal cycle. The spring peak occurs on $100\,\unit{doy}$ and amounts to $40\,\unit{ppb}$ while the annual average \chem{\left<[O_3]\right>} is $28\,\unit{ppb}$. 
The early decline in \chem{\left<[O_3]\right>} may point to an influence of coniferous vegetation on the removal of ozone. Coniferous trees have been observed \chem{CO_2} uptake as early as \unit{doy} 100 \citep{TB:Kolari2007, TP:Wallin2013}. The estimated average begin of growing season for deciduous trees ($\left<G_\mathrm{start}\right>_\mathrm{decid}$) is $(150\pm 14)\,\unit{doy}$ at Svanvik (see Section~\ref{subsec:do3se_parameters} for details). In July--September, ozone is occasionally almost completely depleted during nighttime. This hints to ozone uptake by vegetation even at low light intensities during midnight sun conditions in combination with stable planetary boundary layer conditions preventing mixing of ozone rich air.

\begin{figure}[t]
  %\includegraphics[width=8.3cm]{ozone_climatology_fenoscandic_obs_spline}
  \includegraphics[width=8.3cm]{fig05}
  \caption{Svanhovd ozone climatology. The density distribution is shown together with a multi-annual mean of daily \chem{[O_3]} (climatology). A spline has been fitted through daily mean \chem{[O_3]}. The climatology of daily mean ozone is shown together with $1\,\sigma$ standard deviation and standard error of the mean.}
  \label{fig:ozone_climatology_fenoscandic_obs_spline}
\end{figure}

\subsection{2018/19 anomalies}
\label{subsec:anomalies}
In the following, we look at the 2018/19 anomalies and discuss the differences between the two years. As weather extremes like in 2018 are likely to become more frequent with climate change, this will help to assess possible future changes in the vulnerability of subarctic vegetation to ozone.\\

In Fig.~\ref{fig:anomalies_svanvik}a, temperature anomalies for Svanvik are shown as a percentage of days warmer or colder than $\pm 1\,\sigma$ from the climatology for each month. Negative deviations from climatology are displayed as a negative percentage. The annual positive/negative deviations are indicated in the respective corners (right upper/right lower). The hatched area between the dashed lines indicates the expected percentage of values falling above/below $\pm 1\,\sigma$ if a normal distribution is assumed ($15.9\,\unit{\%}$). We find, that the summer of 2018 was significantly warmer than average. Especially in May and July more than $40\,\unit{\%}$ of the days were significantly warmer. Significantly more warmer days continued to occur in all months (August--November) except for October. March 2018 had many unusually cold days. The summer of 2019 was fairly average, however, significantly more colder days occurred in July, while August had both significantly colder and warmer days. April 2019 had significantly warmer than average days.

@@

Similarly, the precipitation anomalies for Svanvik are displayed in Fig.~\ref{fig:anomalies_svanvik}b. Due to the large interannual variability in observed precipitation, only a few days deviated significantly from the climatology on the $1 \sigma$ level. Therefore, we lowered the $\sigma$ level constraint to $\pm \frac{1}{2}\,\sigma$. The percentage of days wetter/drier than $\pm \frac{1}{2}\,\sigma$ are shown for each month. Positive deviation refers to days wetter than average and negative deviation to drier than average. The hatched area between the dashed lines indicates the expected percentage of values falling above/below $\pm\frac{1}{2}\,\sigma$ if a normal distribution is assumed ($30.8\,\unit{\%}$). Unlike the temperature anomalies, the picture is not as clear. While March 2018 had a significant number of days wetter than usual, summer and fall (May/July, September/October) had a significant number of days drier than average. 2019 was rather average throughout the year, but with a significant number of days drier than normal in September/October and wetter in December.

The global irradiance anomalies for Svanvik are presented in Fig.~\ref{fig:anomalies_svanvik}c. The percentage of days above or below $\pm 1\,\sigma$ from the climatological mean is displayed for each month. In summer 2018 (May/July), global irradiance showed a significant number of days with higher than average irradiance, while 2019 was generally an average year with a significant number of days darker in spring (April/May).

%Ozone
We adjusted the ozone climatology for Svanvik \citep{ACPD:Falk2021} and assess the monthly significance of the ozone concentration anomalies in 2018/19. The resulting percentage of days above/below $\pm 1\,\sigma$ from the bias-corrected climatological mean is displayed for each month in Fig.~\ref{fig:anomalies_svanvik}d. Negative deviations from climatology are shown as a negative percentage. Note that missing bars for certain months are both due to data availability and anomalies not meeting the $1\sigma$ criteria. The star indicates results after the reconstruction of missing data in July. In April/May 2018, ozone was significantly enhanced, while no significant enhancement is found in 2019. In both years, ozone concentrations were significantly lower than average in November. These results indicate that the extensive forest fires in Sweden did not contribute to a significant enhancement in \chem{[O_3]} at Svanhovd in 2018, but the stagnant meteorological conditions in April/May in combination with a late start of the growing season.

In summary, 2018 had a significant number of days that were warmer, drier, and brighter than the climatological average, while 2019 was a rather average year. Ozone was not significantly enhanced during the growing season in 2018, but a substantially higher number of peak \chem{[O_3]} were observed in 2018 than in 2019.
